---
title: "Dallas Vehicle Theft Rates by Year"
author: "Greg Heartsfield"
date: "December 13, 2014"
output: html_document
---

# Introduction

This is a quick attempt to understand theft rates of vehicles in Dallas, based on the [Dallas Open Data](https://www.dallasopendata.com/) website and a very simple exponential model of vehicle life.  This was created during the [Dallas R Meetup Data Hackathon](http://www.meetup.com/Dallas-R-Users-Group/events/218651376/).  The main question I wanted to answer was, do vehicle theft rates change based on vehicle age?  And if so, does this mean that very new cars are more or less likely to be stolen?

# Importing and Exploration

Import libraries for graphing and data manipulation.

```{r warning=FALSE,message=FALSE}
require(lattice)
require(ggplot2)
require(plyr)
```


Read vehicle data from the Dallas Open Data website, if it doesn't already exist in the current directory.

```{r}
vehicle.file <- "rms_vehicle.csv"
if(!file.exists(vehicle.file)) {
  download.file("http://www.dallasopendata.com/api/views/imkj-2sfb/rows.csv?accessType=DOWNLOAD",vehicle.file);
}

```

Read in as CSV.

```{r}
vehicle <- read.csv(vehicle.file,stringsAsFactors=FALSE)
```

Some of the vehicle year data has incorrect years with transposed digits or ambiguous single digits.

```{r}
vehicle[which(vehicle$VehYr >1000 & vehicle$VehYr <1900),]$VehYr
```

So, we only look at years that make sense for vehicle (1975 to 2015).

```{r}
vehicle.correct.years <- vehicle[which((vehicle$VehYr > 1975) &(vehicle$VehYr < 2015)),]
```


There are several reasons for vehicles showing up in the dataset:

```{r}
unique(vehicle$VehType)
```

For our purposes, we will only consider the stolen vehicles.

```{r}
vehicles.stolen <- vehicle.correct.years[which(vehicle$VehType=="Stolen"),]
```

Dotchart showing total number of vehicles stolen by vehicle year.

```{r fig.height=8}
yrs.table <- table(vehicles.stolen$VehYr)
dotchart(as.numeric(yrs.table), labels=row.names(yrs.table))
```

Lets also compare the distributions of the other kinds of vehicle incidents to understand if there is something distinct about thefts.  I've removed some of the types that have very little data.

```{r fig.height=10, fig.width=8}
vehicles.by.type <- data.frame(year=vehicle.correct.years$VehYr, type=vehicle.correct.years$VehType)
vehicle.type.table <- as.data.frame(table(vehicles.by.type))
veh.type.subset <- vehicle.type.table[which(vehicle.type.table$type %in% c("Victim", "Stolen", "Suspect", "Recovered")),]
dotplot(year ~ Freq|type, data=veh.type.subset)
```

The average age of vehicles on the road is 11.4 years (http://www.latimes.com/business/autos/la-fi-hy-ihs-automotive-average-age-car-20140609-story.html).  It is reasonable to assume that a car's "survival" rate may be modeled by an exponential distribution (a car may have a 5% chance of being removed from the population each year).  This isn't perfect, but it's not totally unrealistic.  Based on that, we can determine a very rough estimate of the underlying distribution of car ages.
 
 * Assumptions: number of cars sold each year is constant.
 * Car lifespan is 11.4 years on average.
 * Distribution of car lifespan can be modeled by an exponential distribution.
 * Each make/model of car can be represented by the same distribution.
 
Determine the total number of stolen cars, which will be used to scale the "expected" distribution later.

```{r}
car.stolen.count <- dim(vehicles.stolen)[1]
```

Find a parameter for the exponential function that comes close to 11.4 years.  Use optimize to find a value that minimizes the error by fitting a rate for the exponential distribution.

```{r}
exp.rate.err <- function(r) {
  abs(sum(dexp(seq(1,100), rate=r) * seq(1,100)) - 11.4)
}
exp.rate <- optimize(f=exp.rate.err, interval=c(0,1))$minimum
```

In this case, we estimate the rate as `r exp.rate`.

Look at the expected number of stolen vehicles based on an exponential distribution.

```{r fig.height=8}
exp.stolen.count <- dexp(seq(40,1), rate=exp.rate) * car.stolen.count
dotchart(exp.stolen.count, labels=(as.character(seq(1975,2014))))
```

Create a dataframe combining the expected and actual numbers of stolen vehicles by year.

```{r}
year.range <- seq(1975,2014)
exp.stolen <- data.frame(year = year.range, stolen.count = exp.stolen.count, type="expected")
act.stolen <- data.frame(year=row.names(yrs.table), stolen.count = as.numeric(yrs.table),type="actual")
stolen <- rbind(exp.stolen, act.stolen)
```

# Comparing Actual and Expected

Now compare the expected and actual stolen counts for each year:

```{r}
dotplot(year ~ stolen.count|type, data=stolen)
```

Use ggplot2 for an alternate view that is easier to compare.  This shows the lower-than-expected theft rate for very new cars and very old cars, and a greater-than-expected theft rate for slightly older cars (1999-2008).

```{r fig.height=8, fig.width=8}
ggplot(stolen, aes(x = stolen.count, y = year, fill=type)) +
  geom_dotplot(binaxis = "y") +
  labs(title="Expected vs. Actual Vehicle Thefts by Model Year",x="Vehicles Stolen")
```

There are many potential explanations for this result:

* Car production was not uniform year-to-year, or other defects in the "expected" distribution.
* Very new cars have better anti-theft mechanisms.
* Very new cars less likely to be left in high-theft areas.
* Very old cars not worth the potential trouble to steal (low payoff).

# Theft Risk Ratio

Another way to view this is by the ratio of actual thefts to expected thefts based on our population estimate.

We start by building a new dataframe for each of the years with the actual and expected thefts together.  We'll just omit years with no actual thefts.

```{r}
rr.left <- data.frame(year=seq(1975,2014), expected=exp.stolen$stolen.count)
rr <- merge(rr.left,act.stolen)
rr$type <- NULL
rr <- rename(rr,c("stolen.count"="actual"))
```

Compute the ratio:

```{r}
rr$ratio <- rr$actual / rr$expected
```

```{r fig.height=8, fig.width=8}
dotplot(as.character(year) ~ ratio, data=rr)
```