---
title: "DallasOpenData_Vehicle"
author: "Greg Heartsfield"
date: "December 13, 2014"
output: html_document
---

Import libraries for graphing.

```{r}
require(lattice)
require(ggplot2)
```


Read vehicle data from the Dallas Open Data website, if it doesn't already exist in the current directory.

```{r}
vehicle.file <- "rms_vehicle.csv"
if(!file.exists(vehicle.file)) {
  download.file("http://www.dallasopendata.com/api/views/imkj-2sfb/rows.csv?accessType=DOWNLOAD",vehicle.file);
}

```

Read in as CSV.

```{r}
vehicle <- read.csv(vehicle.file,stringsAsFactors=FALSE)
```

Some of the vehicle year data has incorrect years with transposed digits, so we only look at years that make sense for vehicle (1975 to 2015).

```{r}
vehicle.correct.years <- vehicle[which((vehicle$VehYr > 1975) &(vehicle$VehYr < 2015)),]
```

Look at just stolen vehicles


```{r}
vehicles.stolen <- vehicle.correct.years[which(vehicle$VehType=="Stolen"),]
```

Dotchart showing number of vehicles stolen by vehicle year.

```{r fig.height=8}
yrs.table <- table(vehicles.stolen$VehYr)
dotchart(as.numeric(yrs.table), labels=row.names(yrs.table))
```

Average age of vehicles on road is 11.4 years (http://www.latimes.com/business/autos/la-fi-hy-ihs-automotive-average-age-car-20140609-story.html).  So, based on that, we can determine a very rough estimate of what we would expect the underlying distribution of car ages to be.
 
 * Assumptions: number of cars sold each year is constant.
 * Car lifespan is 11.4 years on average.
 * Distribution of car lifespan is exponential
 
Determine the total number of stolen cars:

```{r}
car.stolen.count <- dim(vehicles.stolen)[1]
```

Find a parameter for the exponential function that comes close to 11.4 years (looking at 50 years of vehicles)
```{r}
sum(dexp(seq(1,50), rate=0.08) * seq(1,50))
```

Look at the expected number of stolen vehicles based on an exponential distribution.

```{r fig.height=8}
exp.stolen.count <- dexp(seq(40,1), rate=0.08) * car.stolen.count
dotchart(exp.stolen.count, labels=(as.character(seq(1975,2014))))
```


Create a dataframe combining the expected and actual numbers of stolen vehicles by year.
```{r}
year.range <- seq(1975,2014)
exp.stolen <- data.frame(year = year.range, stolen.count = exp.stolen.count, type="expected")
act.stolen <- data.frame(year=row.names(yrs.table), stolen.count = as.numeric(yrs.table),type="actual")
stolen <- rbind(exp.stolen, act.stolen)
```

Now compare the expected and actual stolen counts for each year:
```{r}
dotplot(year ~ stolen.count|type, data=stolen)
```

Use ggplot2 for an alternate view

```{r fig.height=8}
ggplot(stolen, aes(x = stolen.count, y = year, fill=type)) +
  geom_dotplot(binaxis = "y") +
  labs(title="Expected vs. Actual Vehicle Thefts by Model Year",x="Vehicles Stolen")
```

